String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-05-05 2023-1-19/sections/introduction.tex
overview

Nevertheless, the aforementioned state-of-the-art string constraint solvers still suffer from the string constraints with regex-counting and string-length, especially when the counting bounds are large. For instance, none of string solvers e.g. CVC4/5, Z3, Z3str3RE, OSTRICH is capable of solving the following constraint within 300 seconds,
%
\begin{equation}\label{eqn-running}
x \in (\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c^+ \wedge |x| > 120.
\end{equation}
Intuitively, the constraint in (\ref{eqn-running}) specifies that $x$ is a concatenation of three subwords $x_1, x_2, x_3$ where $a$ (resp. $b, c$) does not occur in $x_1$ (resp. $x_2, x_3$), moreover, $x$ ends with a nonempty sequence of $c$'s, and the length of $x$ is greater than $120$. It is easy to observe that this constraint is unsatisfiable since on the one hand, $|x| > 120$ and the counting upper bound $60$ in  both $(\Sigma \setminus a)^{\{5, 60\}}$ and $(\Sigma \setminus b)^{\{5, 60\}}$ imply that $x$ must end with some character from $\Sigma \setminus c$, that is, a character different from $c$, and on the other hand, $x \in \Sigma^*c^+$ requires that $x$ has to end with $c$.

A typical way of string constraint solvers to deal with regex-counting is to unfold regular expressions with counting into those \emph{without} counting by utilizing the concatenation operator. For instance, $a^{\{0, 4\}}$ is unfolded into $\varepsilon + a + aa + aaa + aaaa$. Since the unfolding incurs an exponential blow-up on the constraint sizes (where the integer constants are encoded in binary), the unfolding penalizes the performance of the solvers, especially when the bounds are large.

>>>>

Nevertheless, the aforementioned state-of-the-art string constraint solvers still suffer from the string constraints with regex-counting and string-length, especially when the counting bounds are large. For instance, none of string solvers e.g. CVC4/5, Z3, Z3str3RE, OSTRICH is capable of solving the following constraint within 300 seconds,
%
\begin{equation}\label{eqn-running}
x \in (\Sigma \setminus a)^{\{1, 60\}} (\Sigma \setminus b)^{\{1, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c^+ \wedge |x| > 120.
\end{equation}
Intuitively, the constraint in (\ref{eqn-running}) specifies that $x$ is a concatenation of three subwords $x_1, x_2, x_3$ where $a$ (resp. $b, c$) does not occur in $x_1$ (resp. $x_2, x_3$), moreover, $x$ ends with a nonempty sequence of $c$'s, and the length of $x$ is greater than $120$. It is easy to observe that this constraint is unsatisfiable since on the one hand, $|x| > 120$ and the counting upper bound $60$ in  both $(\Sigma \setminus a)^{\{1, 60\}}$ and $(\Sigma \setminus b)^{\{1, 60\}}$ imply that $x$ must end with some character from $\Sigma \setminus c$, that is, a character different from $c$, and on the other hand, $x \in \Sigma^*c^+$ requires that $x$ has to end with $c$.

A typical way of string constraint solvers to deal with regex-counting is to unfold regular expressions with counting into those \emph{without} counting by utilizing the concatenation operator. For instance, $a^{\{1, 4\}}$ is unfolded into $a(\varepsilon + a + aa + aaa)$. Since the unfolding incurs an exponential blow-up on the constraint sizes (where the integer constants are encoded in binary), the unfolding penalizes the performance of the solvers, especially when the bounds are large.

### 2023-05-05 2023-1-19/sections/overview.tex
overview

@@

We first construct an automaton for the regular expression $(\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$. We introduce three registers, say $r_1, r_2, r_3$, to represent the three counting operators.  
%Moreover, we introduce another register, say $r_0$, to denote the length of strings. 
Then we first construct a nondeterministic finite automaton (NFA) for $(\Sigma \setminus a)^* (\Sigma \setminus b)^* (\Sigma \setminus c)^*$, then add the updates of registers to transitions, resulting into a so-called cost-enriched finite automaton (CEFA) $\aut_1$,  as illustrated in Figure~\ref{fig:overview}(a). 
%Then we obtain a cost-enriched finite automaton (CEFA) $\aut_1$ that is illustrated in Figure~\ref{fig:overview}(a), where the self-loops around $p_0, p_1, p_2$ represent the first, second, and third counting operator respectively, in which $r_1, r_2, r_3$ are incremented. 
%Moreover, $r_0$ is incremented in all the three self-loops. 
Furthermore, the counting bounds are specified by the accepting condition $5 \le r_1 \le 60 \wedge 5 \le r_2 \le 60 \wedge 0 \le r_3 \le 60$ that is attached to the final state $p_2$. Note here we represent the counting operators symbolically by registers, instead of unfolding them explicitly. Moreover, we also construct a CEFA $\aut_2$ for $\Sigma^* c^+$ (see Figure~\ref{fig:overview}(b)), which is actually an NFA, since $\Sigma^* c^+$ contains no (explicit) counting operators. 
%where the register $r_0$ is used to record the length of strings.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width = 0.9\textwidth]{sections/overview-cefa.pdf}
  \caption{CEFA for $(\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$ and $\Sigma^* c^+$}
  \label{fig:overview}
\end{figure}

Next, 
%we remove the $\varepsilon$-transitions of $\aut_1$, resulting into $\aut'_1$, 
we construct $\aut_1 \times \aut_2$, that is, the product of $\aut_1$ and $\aut_2$, as illustrated in Figure~\ref{fig:overview:product}(a). Furthermore, we add one special register, say $r_0$, to record the length of strings. Note that $r_0$ is incremented in all transitions. In $\aut_1 \times \aut_2$, the accepting condition $5 \le r_1 \le 60 \wedge 5 \le r_2 \le 60 \wedge 0 \le r_3 \le 60 \wedge r_0 > 120$ is attached to both $(p_0,q_1)$ and $(p_1,q_1)$, where $r_0 > 120$ is added as a conjunct to express $|x| > 120$.  For technical convenience, we also think the updates on the values of registers in transitions as vectors $(v_1, v_2, v_3, v_4)$, where $v_i \in \Int$ is the update on the register $r_{i-1}$ for each $i \in [4]$. For instance, the transitions corresponding to the self-loop around $(p_0, q_0)$ are thought as $(q, a', q', (1,1,0,0))$ with $a' \in \Sigma \setminus \{a\}$, since $r_0$ and $r_1$ are incremented in these transitions. With the updates thought as vectors, the resulting CEFA is illustrated in Figure~\ref{fig:overview:product}(b).
%the transitions corresponding to the self-loop is $(q, a', q', (1,1,0,0))$ with $a' \in \Sigma \setminus \{a\}$.

\begin{figure}[ht]
  \centering
  \includegraphics[width = 0.9\textwidth]{sections/overview-cefa-product.pdf}
  \caption{Product of $\aut_1$ and $\aut_2$, where the length register $r_0$ is added}
  \label{fig:overview:product}
\end{figure}

Let $\tau_1, \cdots, \tau_m$ be an enumeration of the transitions of $\aut_1 \times \aut_2$ and $\myvec{v_1}, \cdots, \myvec{v_m}$ be the vectors of updates on the register values in these transitions. For instance, in the transition from $(p_0, q_1)$ to $(p_1, q_1)$, $r_0$ is incremented, and the values of the other registers is unchanged. Therefore, the vector for this transition is $(1, 0, 0, 0)$. 
Then the satisfiability of the original string constraint is reduced to the nonemptiness of the CEFA $\aut_1 \times \aut_2$. 
From the results in \cite{SSMH04,VSS05}, an existential LIA formula $\psi(i_1, \cdots, i_m)$ can be computed to express the constraints on the number of occurrences of $\tau_1, \cdots, \tau_m$ in the accepting runs of $\aut_1 \times \aut_2$, where $i_1, \cdots, i_m$ are the integer variables representing the number of occurrences of $\tau_1, \cdots, \tau_m$ respectively. 
Therefore, the satisfiability of the string constraint in~(\ref{eqn-running}) is reduced to the satisfiability of the following existential LIA formula
\begin{equation}\label{eqn-LIA}
\psi(i_1, \cdots, i_m) \wedge \bigwedge \limits_{0 \le j \le 3} r_j = \sum \limits_{k \in [m]}  i_k v_{k, j+1} \wedge \alpha(r_0, r_1, r_2, r_3),
\end{equation}
where for each $k \in [m]$, $v_k = (v_{k,1}, v_{k,2}, v_{k, 3}, v_{k, 4})$ is the vector of updates on the registers $r_0, r_1, r_2, r_3$ in the transition $\tau_k$, and $\alpha(r_0, r_1, r_2, r_3)$ is the accepting condition in $\aut_1 \times \aut_2$. Finally, the satisfiability of LIA formula can be solved by the off-the-shelf SMT solvers for LIA.

>>>>

\zhilin{to be made consistent with section 6.}

@@

We first construct an automaton for the regular expression $(\Sigma \setminus a)^{\{1, 60\}} (\Sigma \setminus b)^{\{1, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$. We introduce three registers, say $r_1, r_2, r_3$, to represent the three counting operators.  
%Moreover, we introduce another register, say $r_0$, to denote the length of strings. 
Then we first construct a nondeterministic finite automaton (NFA) for $(\Sigma \setminus a)^* (\Sigma \setminus b)^* (\Sigma \setminus c)^*$, then add the updates of registers to transitions, resulting into a so-called cost-enriched finite automaton (CEFA, see Section~\ref{sec:automaton} for the definition) $\aut_1$,  as illustrated in Figure~\ref{fig:overview}(a). 
%Then we obtain a cost-enriched finite automaton (CEFA) $\aut_1$ that is illustrated in Figure~\ref{fig:overview}(a), where the self-loops around $p_0, p_1, p_2$ represent the first, second, and third counting operator respectively, in which $r_1, r_2, r_3$ are incremented. 
%Moreover, $r_0$ is incremented in all the three self-loops. 
Furthermore, the counting bounds are specified by the accepting condition $1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60$. Note here we represent the counting operators symbolically by registers, instead of unfolding them explicitly. Moreover, we construct another CEFA $\aut_2$ for $\Sigma^* c^+$ (see Figure~\ref{fig:overview}(b)), which is actually an NFA, since $\Sigma^* c^+$ contains no (explicit) counting operators. Finally, we construct the third CEFA $\aut_3$ (see Figure~\ref{fig:overview}(c)), where a register $r_4$ is used to denote the length of strings and the accepting condition is $\ltrue$. 
%where the register $r_0$ is used to record the length of strings.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width = 0.95\textwidth]{sections/overview-cefa.pdf}
  \caption{CEFA for $(\Sigma \setminus a)^{\{1, 60\}} (\Sigma \setminus b)^{\{1, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$, $\Sigma^* c^+$, and $|x|$}
  \label{fig:overview}
\end{figure}

Next, 
%we remove the $\varepsilon$-transitions of $\aut_1$, resulting into $\aut'_1$, 
we construct $\aut_1 \cap \aut_2 \cap \aut_3$, that is, the intersection (aka product) of $\aut_1$, $\aut_2$, and $\aut_3$, as illustrated in Figure~\ref{fig:overview:product}(a), where the states from which no final states are reachable are removed. 
%Furthermore, we add one special register, say $r_0$, to record the length of strings. Note that $r_0$ is incremented in all transitions. 
%In $\aut_1 \cap \aut_2 \cap \aut_3$, the accepting condition $1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60 \wedge r_0 > 120$ is attached to both $(p_0,q_1)$ and $(p_1,q_1)$, where $r_0 > 120$ is added as a conjunct to express $|x| > 120$.  
For technical convenience, we also think the updates on the values of registers in transitions as vectors $(v_1, v_2, v_3, v_4)$, where $v_i \in \Int$ is the update on the register $r_{i}$ for each $i \in [4]$. For instance, the transitions corresponding to the self-loop around $(p_0, q_0, q'_0)$ are thought as $((p_0, q_0, q'_0), a', (p_0, q_0, q'_0), (1,0,0,1))$ with $a' \in \Sigma \setminus \{a\}$, since $r_1$ and $r_4$ are incremented in these transitions. With the updates thought as vectors, the resulting CEFA is illustrated in Figure~\ref{fig:overview:product}(b).
%the transitions corresponding to the self-loop is $(q, a', q', (1,1,0,0))$ with $a' \in \Sigma \setminus \{a\}$.

\begin{figure}[ht]
  \centering
  \includegraphics[width = 0.95\textwidth]{sections/overview-cefa-product.pdf}
  \caption{$\aut_1 \cap \aut_2 \cap \aut_3$: Intersection of $\aut_1$, $\aut_2$, and $\aut_3$}
  \label{fig:overview:product}
\end{figure}

\zhilin{stopped here}

Let $\tau_1, \cdots, \tau_m$ be an enumeration of the transitions of $\aut_1 \cap \aut_2 \cap \aut_3$ and $\myvec{v_1}, \cdots, \myvec{v_m}$ be the vectors of updates on the registers in these transitions. 
%For instance, in the transition from $(p_0, q_1)$ to $(p_1, q_1)$, $r_0$ is incremented, and the values of the other registers is unchanged. Therefore, the vector for this transition is $(1, 0, 0, 0)$. 
Then the satisfiability of the original string constraint is reduced to the nonemptiness of the CEFA $\aut_1 \cap \aut_2 \cap \aut_3$ with respect to the LIA formula $\varphi \equiv r_4 > 120$. 
From the results in \cite{SSMH04,VSS05}, an existential LIA formula $\psi(i_1, \cdots, i_m)$ can be computed to express the constraints on the number of occurrences of $\tau_1, \cdots, \tau_m$ in the accepting runs of $\aut_1 \cap \aut_2 \cap \aut_3$, where $i_1, \cdots, i_m$ are the integer variables representing the number of occurrences of $\tau_1, \cdots, \tau_m$ respectively. 
Therefore, the satisfiability of the string constraint in~(\ref{eqn-running}) is reduced to the satisfiability of the following existential LIA formula
\begin{equation}\label{eqn-LIA}
\psi(i_1, \cdots, i_m) \wedge \bigwedge \limits_{0 \le j \le 3} r_j = \sum \limits_{k \in [m]}  i_k v_{k, j+1} \wedge \alpha(r_0, r_1, r_2, r_3),
\end{equation}
where for each $k \in [m]$, $v_k = (v_{k,1}, v_{k,2}, v_{k, 3}, v_{k, 4})$ is the vector of updates on the registers $r_0, r_1, r_2, r_3$ in the transition $\tau_k$, and $\alpha(r_0, r_1, r_2, r_3)$ is the accepting condition in $\aut_1 \times \aut_2$. Finally, the satisfiability of LIA formula can be solved by the off-the-shelf SMT solvers for LIA.
